---
title: "No Pot of Gold at the End of the Rainbow"
output: html_document
date: "June 5, 2016"
subtitle: Can we forecast gold prices using classical time series models?
---

This work project was developed by Dário Soares (m2015315@novaims.unl.pt), Francisco Lourenço (m2015187@novaims.unl.pt), Hugo Ricardo (m2015343@novaims.unl.pt) and Ricardo Dias (m2015058@novaims.unl.pt)

# Background

```{r}

```

# Methodology

```{r, comment=NA }

```

```{r, message=F, echo=F, warning=F}
# load libraries
library(fGarch)
library(tseries)
library(forecast)
library(fpp)
library(ggplot2)
```

# Data

Describe the data...

```{r, comment=NA }
# Load data
gold <- read.csv("project_gold_gld_daily.txt",header=F, sep=",", as.is=TRUE, skip=54, col.names = c("Date", "Value"))

# Set as time-series
gold.ts0 <- ts(gold$Value)

# Remove last 6 observations
gold.ts <- head(gold.ts0,-6)
length(gold.ts0)
length(gold.ts)
```

# Exploratory Analysis

## Stationarity

We started by plotting the original time series of Gold ETF prices in USD.

```{r, comment=NA }
plot.ts(gold.ts, main="Gold ETF Prices (Daily)")

```

The time series *is not* weakly stationary as the level and variance change over time. Because of this we cannot apply directly the Box-Jenkins methodology, so we will evaluate if first differences are sufficient to turn the time series stationary.

```{r, comment=NA }
gold.dif <- diff(gold.ts)
plot.ts(gold.dif, main="Gold ETF Prices (Daily Differences)")
```

Although the result can be considered adequate concerning the level stationarity, issues regarding [MORE] variance remain. Therefore, we will apply a log transformation [MORE] and a subsequent first difference. 

```{r, comment=NA }
gold.dif.log <- diff(log(gold.ts))
plot.ts(gold.dif.log, main="Gold ETF Prices (Daily Differences of Log(Prices))")
```

Unfortunately the log transformation did not overcome the variance issues. Later on we will try to fit a GARCH model. For now, we will follow a pragmatic approach: we will try to fit an ARIMA  model nonetheless and evaluate at the end the temporal structure of the residuals.

## Seasonality

Theoretically seasonality should not be presented in a financial time series as it would mean an arbritrage edge. 

## ARIMA

```{r, comment=NA }
tsdisplay(gold.dif.log)
Acf(gold.dif.log, plot = FALSE)
Pacf(gold.dif.log, plot = FALSE)
```


```{r, comment=NA }
gold.log <- log(gold.ts)
auto.arima(gold.log)
# ARIMA(0,1,0) with drift  
# Random walk 

```


```{r}
Box.test(gold.dif.log, lag = 20, type = "Ljung-Box")
```

p-value = 0.4783
 There is no well defined temporal structure in the transformed data(gold.dif.log)
 The auto.arima points to a random walk. The Ljung-Box test
 applied to the transformed data does not reject the 
 null, with a p-value of 0.4783. 
 Given these assertions we will fit a linear model to the transformed data
 with only the intercept in order to impose a zero mean series. Then we will
 analyse the ACF and PACF of the linear model residuals and decide for
 a GARCH case.
 
 
```{r}
fitlin <- Arima(gold.dif.log, order = c(0,0,0))
summary(fitlin)
```

The intercept is statisticaly significant for a 5% error type I.
We will try to fit a Arch/Garch as it follows

log(yt) - log(yt-1) = beta0 + epsilon

or

Yt/yt-1 = e^(beta0 + epsilon)

(nota: epsilon = arga)


# ARCH-GARCH


```{r}
arga <- fitlin$residuals
tsdisplay(arga, lag.max = 160)
tsdisplay((arga)**2, lag.max = 160)
```

From the analysis of the ACF and PACF of the squares of the linear model residuals (arga) we conclude that there is a case for a changing variance over time. As the pattern is conducive to a ARMA model we will do a simulation with multiple GARCH models to assist us on the fit decision.

| Model | AIC | Significance
|------|-----|-----|
| GARCH | 12 | 3 |
| GARCH(1,1) | -6.096633  | YES | 
| GARCH(1,0) | -5.944167 | ? | 
| GARCH(2,0) | -5.971479 | ? | 
| GARCH(3,0) | -5.996618 | ? | 
| GARCH(4,0) | -6.014606 | ? | 
| GARCH(5,0) | -6.025685 | ? | 
| GARCH(6,0) | -6.050757 | ? | 
| GARCH(2,1) | -6.095755 | ? | 
| GARCH(3,1) | -6.094847  | ? | 
| GARCH(4,1) | -6.093946  | ? | 
| GARCH(5,1) | -6.093174  | ? | 
| GARCH(6,1) | -6.092338 | ? | 
| GARCH(1,2) | -6.095783  | ? | 
| GARCH(1,3) | -6.094890 | ? | 
| GARCH(1,4) | -6.094029  | ? | 
| GARCH(1,5) | -6.093382 | ? | 
| GARCH(1,6) | -6.092544 | ? | 
| GARCH(2,2) | -6.096449  | ? | 
| GARCH(2,3) | -6.095632 | ? | 
| GARCH(2,4) | -6.094740 | ? | 
| GARCH(2,5) | -6.094125 | ? | 
| GARCH(2,6) | -6.093348 | ? | 
| GARCH(3,2) | -6.095585 | ? | 
| GARCH(3,3) | -6.094197 | ? | 
| GARCH(3,4) | -6.093292 | ? | 
| GARCH(3,5) | -6.093531 | ? |
| GARCH(3,6) | -6.092720 | ? |
| GARCH(4,2) | -6.094991 | ? |
| GARCH(4,3) | -6.093292  | ? |
| GARCH(4,4) | -6.093847 | ? |
| GARCH(4,5) |  -6.093599  | ? |
| GARCH(4,6) | -6.092924  | ? |
| GARCH(5,2) | -6.094174  | ? |
| GARCH(5,3) | -6.092448  | ? |
| GARCH(5,4) | -6.093027  | ? |
| GARCH(5,5) | -6.093050 | ? |
| GARCH(5,6) | -6.092220  | ? |
| GARCH(6,2) | -6.093563  | ? |
| GARCH(6,3) | -6.093281 | ? |
| GARCH(6,4) | -6.093205 | ? |
| GARCH(6,5) | -6.093205  | ? |
| GARCH(6,6) | -6.093019  | ? |

From this simulation we concluded the est model was a garch(1,1), according to the parsimony principle

FALTA: especificação do GARCH

```{r}
gfit <- garchFit(~garch(1,1), arga, trace=F)
summary(gfit)
```

The tests results were good excepted for normality. The fit wasn't able to recover and capture the full extreme values dynamics. Anyway we pragmatically will go to forecasting. 

```{r}
u <- gfit@sigma.t
argaplusu <- arga+1.96*u
argaminusu <- arga-1.96*u
plot(arga)
lines(argaplusu, lty= 2,col = 3)
lines(argaminusu, lty= 2,col = 3)
```

```{r}
x <- predict(gfit, n.ahead=6)
str(x)
x
argaforecast <- x$meanForecast
argaforecastplusu <- argaforecast + 1.96*(x$meanError)
argaforecastminusu <- argaforecast - 1.96*(x$meanError)
```

```{r}
# voltar aos valores iniciais
# intercet no modelo fitlin fitlin$coef

# ??? C1
argaI<- arga + fitlin$coef
argaplusuI<-argaplusu + fitlin$coef
argaminusuI<-argaminusu + fitlin$coef

# ??? C2
argaforecastI <- argaforecast + fitlin$coef
argaforecastplusuI<- argaforecastplusu  + fitlin$coef
argaforecastminusuI<- argaforecastminusu  + fitlin$coef

# ??? C3
gold.ts.1 <- head(gold.ts,-1)
gold.ts.arga<- (exp(argaI))*gold.ts.1
gold.ts.argaplusu<- (exp(argaplusuI))*gold.ts.1
gold.ts.argaminusu<- (exp(argaminusuI))*gold.ts.1

# ??? C4
gold.ts.argaforecast1<- (exp(argaforecastI[1]))*gold.ts[length(gold.ts)]
gold.ts.argaforecast2<- (exp(argaforecastI[2]))*gold.ts.argaforecast1
gold.ts.argaforecast3<- (exp(argaforecastI[3]))*gold.ts.argaforecast2
gold.ts.argaforecast4<- (exp(argaforecastI[4]))*gold.ts.argaforecast3
gold.ts.argaforecast5<- (exp(argaforecastI[5]))*gold.ts.argaforecast4
gold.ts.argaforecast6<- (exp(argaforecastI[6]))*gold.ts.argaforecast5
gold.ts.argaforecast <-c(gold.ts.argaforecast1,gold.ts.argaforecast2,gold.ts.argaforecast3,gold.ts.argaforecast4,gold.ts.argaforecast5,gold.ts.argaforecast6)

# ??? C5
gold.ts.argaforecastplusu1<- (exp(argaforecastplusuI[1]))*gold.ts[length(gold.ts)]
gold.ts.argaforecastplusu2<- (exp(argaforecastplusuI[2]))*gold.ts.argaforecastplusu1
gold.ts.argaforecastplusu3<- (exp(argaforecastplusuI[3]))*gold.ts.argaforecastplusu2
gold.ts.argaforecastplusu4<- (exp(argaforecastplusuI[4]))*gold.ts.argaforecastplusu3
gold.ts.argaforecastplusu5<- (exp(argaforecastplusuI[5]))*gold.ts.argaforecastplusu4
gold.ts.argaforecastplusu6<- (exp(argaforecastplusuI[6]))*gold.ts.argaforecastplusu5
gold.ts.argaforecastplusu <- c(gold.ts.argaforecastplusu1, gold.ts.argaforecastplusu2, gold.ts.argaforecastplusu3, gold.ts.argaforecastplusu4, gold.ts.argaforecastplusu5, gold.ts.argaforecastplusu6)

# ??? C6
gold.ts.argaforecastminusu1<- (exp(argaforecastminusuI[1]))*gold.ts[length(gold.ts)]
gold.ts.argaforecastminusu2<- (exp(argaforecastminusuI[2]))*gold.ts.argaforecastminusu1
gold.ts.argaforecastminusu3<- (exp(argaforecastminusuI[3]))*gold.ts.argaforecastminusu2
gold.ts.argaforecastminusu4<- (exp(argaforecastminusuI[4]))*gold.ts.argaforecastminusu3
gold.ts.argaforecastminusu5<- (exp(argaforecastminusuI[5]))*gold.ts.argaforecastminusu4
gold.ts.argaforecastminusu6<- (exp(argaforecastminusuI[6]))*gold.ts.argaforecastminusu5
gold.ts.argaforecastminusu <- c(gold.ts.argaforecastminusu1, gold.ts.argaforecastminusu2, gold.ts.argaforecastminusu3, gold.ts.argaforecastminusu4, gold.ts.argaforecastminusu5, gold.ts.argaforecastminusu6)

#plot(gold.ts)
#lines(gold.ts.arga, lty= 2,col = 3)
#lines(gold.ts.argaplusu, lty= 2,col = 4)
#lines(gold.ts.argaminusu, lty= 2,col = 4)
#lines(c(gold.ts, gold.ts.argaforecast), lty= 2, col = 5)

# prepare data for plots
x.lagged <- tail(gold.ts,-1) #remove first observation
x.series <- tail(x.lagged,6)
x.arga <- tail(gold.ts.arga,6)
x.argaplusu <- tail(gold.ts.argaplusu, 6)
x.argaminusu <- tail(gold.ts.argaminusu, 6)

# prepare dataframe for ggplot2
x.t <- c(1,2,3,4,5,6)
x.bind <- as.data.frame(cbind(x.t, x.series, x.arga, x.argaplusu, x.argaminusu))
x.bind

```

```{r}
# plot forecasts
p <- ggplot(x.bind, aes(x.t)) + 
  geom_line(aes(y=x.arga), colour="blue") + 
  geom_line(aes(y=x.argaplusu), colour="red") + 
  geom_line(aes(y=x.argaminusu), colour="red")
p + theme_bw()  
```



# Exponential Smoothing Forecast

Em baixo tentámos modelar a nossa séries e fazer previsões recorrendo ao alisamento exponencial.
Logo à partida, decidimos não utilizar o método de alisamento exponencial simples, uma vez que a nossa série é caracterizada por ter tendência e não sazonalidade.

Os métodos de alisamento mais indicados serão portanto o Holt's Linear Trend, Holt's Exponential, Damped Linear Trend e,  Damped Exponential Trend.

```{r}
fit1<-holt(gold.ts, h=6)
fit2<-holt(gold.ts, initial='optimal', exponential=TRUE, h=6)
fit3<-holt(gold.ts, initial='optimal', damped=TRUE, h=6)
fit4<-holt(gold.ts, initial='optimal', exponential=TRUE, damped=TRUE, h=6)
```

Segundo o AIC o melhor modelo para a nossa série é o Holt's Exponential.
Este modelo é mais adequado para séries com tendência e sem sazonalidade como é o caso da nossa.
Resta comprovar se o erro de previsão do modelo Holt's Exponential é menor do que o do modelo Damped Exponential Trend cujo AIC foi o segundo mais baixo.

```{r}
par(mfrow=c(2,2))
plot(tail(gold.ts,20), main= paste("Holt's Linear"," AIC=", round(AIC(fit1$model),1)), 
     xlab="", ylab = "Prices", col="black", type="l", cex.main=0.9)
lines(tail(fitted(fit1),20), col="cyan")
plot(tail(gold.ts,20), main= paste("Holt's Exponential"," AIC=", round(AIC(fit2$model),1)), 
     xlab="", ylab = "Prices", col="black", type="l", cex.main=0.9)
lines(tail(fitted(fit2),20), col="blue")
plot(tail(gold.ts,20), main= paste("Damped Trend"," AIC=", round(AIC(fit3$model),1)), 
     xlab="", ylab = "Prices", col="black", type="l", cex.main=0.9)
lines(tail(fitted(fit3),20), col="green")
plot(tail(gold.ts,20), main= paste("Exponential Damped Trend"," AIC=", round(AIC(fit4$model),1)), 
     xlab="", ylab = "Prices", col="black", type="l", cex.main=0.9)
lines(tail(fitted(fit4),20), col="red")
```




```{r}
# plot forecasts
par(mfrow=c(1,1))
plot(fit1$mean, col="cyan", xlab="Days", ylab="Prices", main="Sample Forecasts (next 6 days)")
par(new=T)
plot(fit2$mean, col="blue", xlab="", ylab="", axes=F)
par(new=T)
plot(fit3$mean, col="green", xlab="", ylab="", axes=F)
par(new=T)
plot(fit4$mean, col="red", xlab="", ylab="", axes=F)
legend("topleft", lty=1, col=c("cyan","blue","green","red"),
       c("Holt's Linear", "Holt's Exponential","Damped Trend","Exponential Damped Trend"), cex = 1)
```

# Discussion

# References

TODO: DARIO



